---
title: "STAT 4620 Final Project Report"
author: "Daphne Kaur, Arjun Narang, Charlene Li, and Samhit Kasichainula"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(corrplot)
library(ggplot2)
library(patchwork)
library(mgcv)
library(tree)
library(randomForest)
library(gbm)
library(corrplot)
library(dplyr)
library(knitr)
library(kableExtra)
library(tidyr)

```


```{R echo=FALSE}  
# Data Loading
load("Wage_Stat4620_2023.RData")
data <- Wage_Stat4620
```


# 1. Introduction

Our goal is to model the relationship between workers’ wages and a set of demographic and labor-market characteristics using the Wage dataset from the ISLR package. Understanding how factors such as age, education, job classification, year, and marital status contribute to wage outcomes is important for both labor economics research and practical workforce planning. The objective is to identify a model that is both interpretable and well-supported by the data, and to summarize insights about which variables most strongly influence wages and how these effects behave across demographic groups.


# 2. Data Description and Cleaning

**Dataset Structure**

The Wage dataset includes 3000 observations with 13 variables. Table 1 shows the description and measurement type for each variable.
```{r echo=FALSE, message=FALSE, warning=FALSE}


var_summary <- data.frame(
  Variable = c("X", "year", "age", "maritl", "race", "education", "region",
               "jobclass", "health", "health_ins", "logwage", "wage", "Resp"),
  Description = c("Respondent ID (unique identifier)", 
                  "Year that wage information was recorded", 
                  "Age of worker", 
                  "Marital status: 1. Never Married, 2. Married, 3. Widowed, 4. Divorced, 5. Separated", 
                  "Race of worker: 1. White, 2. Black, 3. Asian,  4. Other", 
                  "Highest education level: 1. < HS Grad, 2. HS Grad, 3. Some College, 4. College Grad, 5. Advanced Degree", 
                  "Region of country (mid-atlantic only)", 
                  "Job class: 1. Industrial, 2. Information", 
                  "Self-reported health status: 1. <=Good, 2. >=Very Good ", 
                  "Health insurance status: 1. Yes, 2. No", 
                  "Log-transformed workers wage", 
                  "Workers raw wage", 
                  "Mystery response variable"),
  Type = c("Numerical: Discrete",
           "Numerical: Discrete", "Numerical: Discrete", "Categorical: Nominal", "Categorical: Nominal",
           "Categorical: Ordinal", "Categorical: Nominal", "Categorical: Nominal", "Categorical: Ordinal",
           "Categorical: Binary", "Numerical: Continuous", "Numerical: Continuous", "Numerical: Continuous")
)

kable(var_summary, caption = "Variable Descriptions for Wage Dataset", booktabs = TRUE, linesep = "") %>%
  kable_styling(full_width = TRUE, font_size = 7) %>%   # make it wider and smaller
  column_spec(2, width = "25em", extra_css = "word-wrap: break-word; white-space: normal;") %>% # wider Description
  column_spec(1, width = "5em") %>%                     # Variable column
  column_spec(3, width = "11em")                        # Type column

```

**Data Cleaning & Preprocessing**

All original variables are complete. The appended variable Resp contains missing values. To examine potential bias, we created an indicator variable marking whether Resp was missing. Comparing summary statistics of numerical predictors between missing and non-missing Resp cases showed similar mean values, suggesting that the missingness is likely missing at random (Table 2). Therefore, removing rows with missing Resp will not bias the analysis. We also checked for duplicate rows and found none. Thus, our final cleaned dataset excludes the identifier variable X and all rows with missing Resp.
```{r echo=FALSE, message=FALSE, warning=FALSE}
data %>%
  mutate(missing = is.na(Resp)) %>%
  group_by(missing) %>%
  summarize(across(where(is.numeric), ~round(mean(.x, na.rm = TRUE), 2))) %>%
  kable(caption = "Mean of Numeric Variables by Missingness of Resp", booktabs = TRUE, linesep = "") %>%
  kable_styling(full_width = FALSE, font_size = 7) %>%  # smaller font for less vertical space
  row_spec(0, extra_css = "line-height: 0.9;") %>%      # header row spacing
  row_spec(1:2, extra_css = "line-height: 0.9;") # data row spacing

```

```{R echo=FALSE, message=FALSE, warning=FALSE, results='hide'} 
# Convert categorical variables to factors
data$maritl <- as.factor(data$maritl)
data$race <- as.factor(data$race)
data$education <- as.factor(data$education)
data$region <- as.factor(data$region)
data$jobclass <- as.factor(data$jobclass)
data$health <- as.factor(data$health)
data$health_ins <- as.factor(data$health_ins)

# Check duplicates
sum(duplicated(data))
```

```{R echo=FALSE, message=FALSE, warning=FALSE} 
data_clean <- data %>%
  dplyr::select(-X) %>%
  dplyr::filter(!is.na(Resp))

```


# 3. Exploratory Data Analysis

## 3.1 Distribution of the Response Variable (Resp)

Before examining predictors, we examine the shape of the response variable. The histogram below shows that **Resp is clearly bimodal**, with two distinct peaks. The strong **bimodal distribution** suggests that a linear mean structure may not fully capture the data, and some predictors may shift individuals into one mode or the other. This motivates the need for polynomial and semi-parametric models.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=1.5}
ggplot(data_clean, aes(x = Resp)) +
  geom_histogram(binwidth = 0.3, fill = "grey", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Resp (Response Variable)",
        x = "Resp", y = "Count")+
  theme(
    plot.title = element_text(size = 8),   # smaller title
    axis.title = element_text(size = 7),    # smaller axis labels
    axis.text = element_text(size = 6)      # smaller axis tick labels
  )
```

## 3.2 Numeric Variables & Categorical Variables

The variable wage shows a wide range, with a median that is lower than the mean, indicating the presence of some high earners. The variable logwage is the logarithmic transformation of wage, which reduces skewness and makes the distribution more symmetric (Table 3). The categorical variables were summarized by their most frequent categories (Table 4).

```{R echo=FALSE, message=FALSE, warning=FALSE} 
# Numeric summary
numeric_summary <- data_clean %>%
  dplyr::select(where(is.numeric)) %>%
  dplyr::summarize(
    dplyr::across(
      dplyr::everything(),
      list(
        Min = ~round(min(.x, na.rm = TRUE), 2),
        Median = ~round(median(.x, na.rm = TRUE), 2),
        Mean = ~round(mean(.x, na.rm = TRUE), 2),
        Max = ~round(max(.x, na.rm = TRUE), 2)
      )
    )
  )

# Reshape for reporting
numeric_summary_t <- numeric_summary %>%
  pivot_longer(
    everything(),
    names_to = c("Variable", "Statistic"),
    names_pattern = "(.*)_(Min|Median|Mean|Max)"
  ) %>%
  pivot_wider(names_from = Statistic, values_from = value)

kable(
  numeric_summary_t,
  caption = "Summary Statistics – Numeric Variables",
  booktabs = TRUE,
  format = "latex"
) %>%
  kable_styling(font_size = 6.5, latex_options = c("hold_position")) %>%
  column_spec(1, width = "1cm") %>%
  column_spec(2:5, width = "1cm")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Categorical (mode) summary
mode_summary <- data_clean %>%
  select(where(is.factor)) %>%
  summarise(across(
    everything(),
    ~ names(sort(table(.x), decreasing = TRUE))[1]
  ))

# Reshape for printing
mode_summary_t <- mode_summary %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Most Frequent Category")

# Print compact table
kable(mode_summary_t, 
      caption = "Most Frequent Category – Categorical Variables", 
      booktabs = TRUE, 
      linesep = "") %>%
  kable_styling(font_size = 7, latex_options = "scale_down") %>%
  row_spec(0, extra_css = "line-height: 0.9;") %>%
  row_spec(1:nrow(mode_summary_t), extra_css = "line-height: 0.9;")
```


**Correlation Analysis (Appendix 1)**

The variables wage and logwage are highly correlated, as expected due to the log transformation. The response variable Resp shows moderate positive correlations with wage, logwage, and age, suggesting that these variables may be influential in predicting Resp. The year variable shows a weak correlation with Resp, indicating limited time-based effects.

**Numeric Variables vs Response (Appendix 2)**

The scatterplot matrix shows a positive, non-linear relationship: wages increase with age up to around 60, then level off or slightly decrease, with variance also growing with age. **Resp is highly correlated with wage and logwage and positively associated with age**. The clusters in the age versus Resp plot likely result from a strong categorical predictor, which can create distinct bands when plotted against a continuous variable. Similarly, the wage and logwage versus Resp plots show clusters influenced by the same categorical predictor.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=2}
# 1 row, 5 columns layout
numeric_vars <- names(data)[sapply(data, is.numeric)]
cat_vars <- c("education", "race", "maritl", "jobclass", "health", "health_ins")
par(mfrow = c(1, 5),
    mar = c(3, 3, 2, 1))  # margins: bottom, left, top, right

# Loop over numeric columns and plot vs Resp
numeric_cols <- colnames(numeric_vars)  # your numeric columns

for(v in numeric_cols) {
  plot(data_clean[[v]], data_clean$Resp,
       pch = 19, col = "darkblue",  # small points
       cex = 0.3,                    # point size
       main = paste(v, "vs Resp"), 
       cex.main = 1,                 # title size
       xlab = v, ylab = "Resp", 
       cex.lab = 0.9,                # axis label size
       cex.axis = 0.7)               # axis tick labels size
}
```

**Categorical Variables vs Response (Appendix 3)**

* education: Clear, strong separation between levels. Higher education corresponds to higher Resp.
* maritl (Married vs Unmarried): Marital status shows one of the strongest relationships in the data, with married individuals disproportionately appearing in the upper mode of the Resp distribution. This pattern likely contributes to the bimodal distribution shape.
* race, health, health_ins: These show weaker group separation. Their effects appear smaller and do not meaningfully explain the bimodality of Resp.
* jobclass: Practically no separation in Resp; likely not an important predictor.
* Outlier Identification: The outliers show data points with unusually low or high wages for their respective categories, which may warrant further investigation for potential influential observations.

## 3.3 Recoding Marital Status (Married vs. Unmarried)

The boxplots below show a clear distinction between the married and unmarried classes of the maritl variable in relationship to the response variable. After coding the maritl variable as binary (1 = Married, 0 = Unmarried), plotting the response and coding it according to the new binary variable, we see that the the two distinct modes in the bimodal response appear to be explained by marital status. Therefore, for the remainder of this report, maritl will be modeled as a binary variable.

```{r, include = FALSE}
# Save original copy of marital status to use for plotting
data_clean$maritl_original <- data_clean$maritl

# Change marital status to married vs not married
data_clean$maritl <- ifelse(data_clean$maritl == "2. Married","Married", "Not Married")
data_clean$maritl <- factor(data_clean$maritl,levels = c("Not Married", "Married"))
```

```{r, echo = FALSE, message = FALSE, fig.align='center', warning = FALSE, out.width= "100%", fig.height = 2}
small_theme <- theme(
  plot.title = element_text(size = 7),
  axis.title = element_text(size = 6),
  axis.text  = element_text(size = 5),
  legend.title = element_text(size = 6),
  legend.text  = element_text(size = 6)
)

p_hist <- ggplot(data_clean, aes(x = Resp, fill = maritl, color = maritl)) +
  geom_histogram(position = "identity", alpha = 0.4, bins = 25) +
  scale_fill_manual(values = c("Not Married" = "#FF9999", "Married" = "#66CCCC")) +
  scale_color_manual(values = c("Not Married" = "#FF7777", "Married" = "#55BBBB")) +
  labs(
    title = "Distribution of Resp by Marital Status",
    x = "Resp",
    y = "Count"
  ) +
  theme_minimal() +
  small_theme

p_box_before <- ggplot(data_clean, aes(x = maritl_original, y = Resp)) +
  geom_boxplot(fill = "#cccccc", outlier.size = 0.5) +
  labs(title = "Before Recoding",
       x = "Original Marital Status",
       y = "Resp") +
  theme_minimal() +
  small_theme+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
)

p_box_after <- ggplot(data_clean, aes(x = maritl, y = Resp)) +
  geom_boxplot(fill = "#66CCCC", outlier.size = 0.5) +
  labs(title = "After Recoding",
       x = "Recoded Marital Status",
       y = "Resp") +
  theme_minimal() +
  small_theme

(p_hist | p_box_before | p_box_after)
```

## 3.4 Understanding Interactions

**Resp vs log(Wage)**

While the model without marital status shows one smooth upward trend, the inclusion of marital status produces two distinct trajectories, especially at higher wage levels, illustrating a meaningful interaction effect.
```{r, echo=FALSE, fig.width=10, fig.height=2, fig.align='center', message=FALSE, warning=FALSE}
# WITHOUT interaction: one smooth curve, ignoring marital status
logwage_maritl_plot <- ggplot(data_clean, aes(x = logwage, y = Resp)) +
  geom_point(alpha = 0.05, size = 0.4) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  labs(title = "Resp vs logwage",
        x = "log(wage)", y = "Resp") +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

# --- Plotting Function (Re-defined for completeness) ---
plot_cat <- function(var) {
  ggplot(data_clean, aes(x = logwage, y = Resp, color = .data[[var]])) +
    geom_point(alpha = 0.05, size = 0.4) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
    labs(title = paste("Resp vs Logwage by", gsub("_", " ", var)),
        x = "log(wage)", y = "Resp") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "right", plot.title = element_text(hjust = 0.5))
}

# --- Data Transformation (Creating the Simplified Variable - FIXED) ---
# Recode based on the numeric factor codes: 2 is 'Married', everything else is 'Not Married'.
# The creation of the new factor is done in one step to avoid the unused argument error.
data_clean <- data_clean %>%
  mutate(
    maritl_simplified = factor(
      case_when(
        as.numeric(maritl_original) == 2 ~ "Married", 
        TRUE ~ "Not Married" 
      )
    )
  )

# --- Original Plot (5-Level Marital Status) ---
# FIX: Changed "maritl" to "maritl_original" to use the 5-level factor
logwage_maritl_plot_int_5lvl <- plot_cat("maritl_original") +
  labs(title = "5-Level Marital Status") +
  guides(color = guide_legend(title = "maritl"))


# --- New Plot (Simplified 2-Level Marital Status) ---
logwage_maritl_plot_int_2lvl <- plot_cat("maritl_simplified") +
  labs(title = "2-Level Marital Status (Married/Not Married)") +
  guides(color = guide_legend(title = "maritl"))


# --- Display Both Charts for Comparison ---
# Compare the original 5-level plot with the new 2-level plot
(logwage_maritl_plot | logwage_maritl_plot_int_5lvl | logwage_maritl_plot_int_2lvl)
```


**Resp vs Age**

Including marital status reveals an interaction effect, where married individuals display consistently higher and more rapidly increasing response values across age.
```{r, echo=FALSE, fig.width=10, fig.height=2, fig.align='center', message=FALSE, warning=FALSE}
# --- Plotting Function (Re-defined for completeness) ---
plot_age_cat <- function(var) {
  ggplot(data_clean, aes(x = age, y = Resp, color = .data[[var]])) +
    geom_point(alpha = 0.05, size = 0.4) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
    labs(title = paste("Resp vs Age by", gsub("_", " ", var)),
         x = "Age", y = "Resp") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "right")
}

# --- 1. Plot WITHOUT Marital Status (Baseline) ---
age_maritl_plot <- ggplot(data_clean, aes(x = age, y = Resp)) +
  geom_point(alpha = 0.05, size = 0.4) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  labs(title = "Resp vs Age (Overall Trend)", x = "Age", y = "Resp") +
  theme_minimal(base_size = 10) + 
  theme( plot.title = element_text(hjust = 0.5))

# --- 2. Plot WITH Original 5-Level Marital Status ---
age_maritl_plot_int_5lvl <- plot_age_cat('maritl_original') +
  labs(title = "Resp vs Age (5-Level Maritl)") +
  guides(color = guide_legend(title = "maritl"))


# --- 3. Plot WITH Simplified 2-Level Marital Status ---
age_maritl_plot_int_2lvl <- plot_age_cat('maritl_simplified') +
  labs(title = "Resp vs Age (2-Level Maritl)") +
  guides(color = guide_legend(title = "maritl"))


# --- Display All Three Charts Side-by-Side ---
# (Overall Trend | 5-Level Comparison | 2-Level Comparison)
(age_maritl_plot | age_maritl_plot_int_5lvl | age_maritl_plot_int_2lvl)
```

# 4. Modeling Strategy and Results

## 4.1 Data Partitioning and Modeling Approach

The cleaned data ($n=2,940$) was partitioned into an **80% training set** ($\text{Wage.train}$) for model fitting and hyperparameter tuning, and a **20% test set** ($\text{Wage.test}$) for unbiased performance evaluation.Our modeling strategy moved from simple parametric models to complex non-parametric methods to find the optimal balance between predictive accuracy (low Test MSE) and statistical inference (interpretability):

* **Interpretable Non-Linear Models:** (Polynomial Regression, GAMs) capture non-linear trends and key interactions.
* **Non-Parametric Ensemble Models:** (Random Forest, Boosting) offer maximum predictive power by capturing arbitrary interactions.

```{r, include= FALSE}
set.seed(4620)
n <- nrow(data_clean)
train_size <- round(n * 0.8) # Create an 80/20 split
train <- sample(1:n,train_size, replace = FALSE)
test <- -train

Wage.train <- data_clean[train, ]
Wage.test  <- data_clean[test, ]
```

## 4.2 Interpretable Non-Linear Models

### 4.2.1 Polynomial Regression:

Polynomial regression was first used as a simple and interpretable way to introduce curvature into the model, capturing **nonlinear patterns** while **preserving a largely linear structure** and serving as a **baseline for assessing nonlinearity**. Models were fit both with and without interaction terms for marital status to evaluate whether these interactions improved accuracy. The best-performing polynomial model used a third-degree polynomial for $\log(\text{wage})$ and a second-degree polynomial for $\text{age}$, with both terms fully interacting with marital status (recoded as Married/Not Married), along with all other categorical predictors.

$$\text{Resp} \sim \text{poly}(\text{logwage}, 3) \times \text{maritl} + \text{poly}(\text{age}, 2) \times \text{maritl} + \text{All Categoricals}$$

```{r, include = FALSE}
# log(Wage)

# Fit 3rd degree polynomial using basic polynomial and with interaction effects. And include GAM.
fit_poly_3 <- lm(Resp ~ poly(logwage, 3), data = Wage.train)
fit_poly3_int <- lm(Resp ~ poly(logwage, 3) * maritl, data = Wage.train)
fit_gam_3 <- gam(Resp ~ s(logwage, by = maritl) + maritl, data = Wage.train)

# Age
# Fit 2nd-degree polynomial using basic, interaction, and GAM for age
fit_age_poly2     <- lm(Resp ~ poly(age, 2), data = Wage.train)
fit_age_poly2_int <- lm(Resp ~ poly(age, 2) * maritl, data = Wage.train)
fit_age_gam       <- gam(Resp ~ s(age, by = maritl) + maritl, data = Wage.train)

# Full model with all categories
poly_full_model <- lm(
  Resp ~ poly(logwage, 3) * maritl + poly(age, 2) * maritl,
  data = Wage.train
)

# Polynomial for logwage AND age (no interaction effects)
poly_both_no_int <- lm(
  Resp ~ poly(logwage, 3) + poly(age, 2),
  data = Wage.train
)
```

```{r, echo = FALSE, warning=FALSE}
# Degree 3 predictions + MSE
pred3 <- predict(fit_poly_3, newdata = data_clean[test, ])
pred3_int <- predict(fit_poly3_int, newdata = data_clean[test, ]) 
pred_gam_3 <- predict(fit_gam_3, newdata = data_clean[test, ])

# Predictions + MSE for age-based models
pred_age2      <- predict(fit_age_poly2,     newdata = Wage.test)
pred_age2_int  <- predict(fit_age_poly2_int, newdata = Wage.test)
pred_age_gam   <- predict(fit_age_gam,       newdata = Wage.test)

# Predict Full Model
pred_poly_full  <- predict(poly_full_model, newdata = Wage.test)
pred_poly_both_no_int <- predict(poly_both_no_int, newdata = Wage.test)

# Calculate the MSE 
mse3  <- mean((pred3 - data_clean[test, ]$Resp)^2)
mse3_int <- mean((pred3_int - data_clean[test, ]$Resp)^2)
mse3_gam <- mean((pred_gam_3 - data_clean[test, ]$Resp)^2)

mse_age2      <- mean((pred_age2     - Wage.test$Resp)^2)
mse_age2_int  <- mean((pred_age2_int - Wage.test$Resp)^2)
mse_age2_gam  <- mean((pred_age_gam  - Wage.test$Resp)^2)

mse_poly_full <- mean((pred_poly_full     - Wage.test$Resp)^2)
mse_poly_both_no_int  <- mean((pred_poly_both_no_int - Wage.test$Resp)^2)

mse_results <- data.frame(
  Model = c("Age Polynomial (Degree 2)",
            "log(Wage) Polynomial (Degree 3)",
            "Age Polynomial (Degree 2) with Marriage Interaction",
            "log(Wage) Polynomial (Degree 3) with Marriage Interaction",
            "Age and log(wage) Polynomial (no interaction)",
            "Age and log(wage) with marital status interaction"),
  Test_MSE = c(mse_age2,
               mse3,
               mse_age2_int,
               mse3_int,
               mse_poly_both_no_int,
               mse_poly_full)
)

kable(mse_results, caption = "log(Wage) and Age Polynomial Test MSE Results") %>%
  kable_styling(full_width = FALSE, font_size = 7)
```

The results show that adding nonlinear terms for both predictors does not improve performance by much unless interaction effects are included, indicating that subgroup differences by marital status drive much of the predictive gain.



```{r, echo = FALSE, warning = FALSE, include= FALSE}
poly_full_all <- lm(
  Resp ~ 
    poly(logwage, 3) * maritl +
    poly(age, 2) * maritl +
    education + race + jobclass + health + health_ins,
  data = Wage.train
)

pred_poly_full_all  <- predict(poly_full_all, newdata = Wage.test)

mse_poly_full_all     <- mean((pred_poly_full_all     - Wage.test$Resp)^2)
mse_full_all_results <- data.frame(
  Model    = c("Age and log(wage) with marital status interaction + all other variables"),
  Test_MSE = c(mse_poly_full_all)
)

kable(mse_full_all_results, caption = "Full Model Polynomial Regression Test MSE Results") %>%
  kable_styling(full_width = FALSE, font_size = 7)

```

**Introducing Categorical Predictors**

In an effort to improve the model's performance, the age and log(Wage) model with interaction effects is set as the baseline. From there, various categorical values were tested to gauge an improvement in Test MSE. The top-performing polynomial models all include the full or near-full set of categorical predictors, suggesting that demographic and job-related characteristics contribute additional explanatory value beyond nonlinear age and wage effects.

```{r, echo = FALSE, warning = FALSE}
# cat_vars: character vector of categorical variables to consider
# train, test: your train/test data frames (e.g., Wage.train, Wage.test)

test_poly_subsets <- function(train, test, cat_vars) {
  
  base_str <- "Resp ~ poly(logwage, 3) * maritl + poly(age, 2) * maritl"
  
  # generate all subsets of cat_vars (including empty set)
  all_subsets <- unlist(
    lapply(0:length(cat_vars), function(k) {
      if (k == 0) {
        list(character(0))   # empty subset
      } else {
        combn(cat_vars, k, simplify = FALSE)
      }
    }),
    recursive = FALSE
  )
  
  results <- data.frame()
  
  for (vars in all_subsets) {
    if (length(vars) == 0) {
      # base model only
      formula_str <- base_str
      label <- "Base: poly(logwage,3)*maritl + poly(age,2)*maritl"
    } else {
      added <- paste(vars, collapse = " + ")
      formula_str <- paste(base_str, "+", added)
      label <- paste0("Base + ", added)
    }
    
    fml <- as.formula(formula_str)
    
    # fit model
    fit <- lm(fml, data = train)
    
    # predict & compute Test MSE
    preds <- predict(fit, newdata = test)
    mse   <- mean((test$Resp - preds)^2)
    
    results <- rbind(
      results,
      data.frame(
        Model        = label,
        Test_MSE     = mse,
        stringsAsFactors = FALSE
      )
    )
  }
  
  # sort by Test MSE (best first)
  results[order(results$Test_MSE), ]
}


cat_vars <- c("education", "race", "jobclass", "health", "health_ins")

poly_subset_results <- test_poly_subsets(
  train   = Wage.train,
  test    = Wage.test,
  cat_vars = cat_vars
)


poly_subset_results %>%
  arrange(Test_MSE) %>%
  slice(1:5) %>%
  kable(
    caption = "Top 5 Polynomial Models by Test MSE",
    digits = 3
  ) %>%
  kable_styling(full_width = FALSE, font_size = 7)

```

### 4.2.2 Generalized Additive Models (GAM):

GAMs were introduced to address the limitations of polynomial models by learning smooth, data-driven functions rather than relying on fixed polynomial shapes, allowing flexible nonlinear modeling while maintaining interpretability. They also permit different smooths by marital status, aligning with previously observed interaction effects, and the model was later expanded to include additional categorical predictors such as education, race, and job class. A GAM was then fitted to replicate the baseline structure of the polynomial models using  $\log(\text{wage})$ and Age with their marital-status interactions. Unlike polynomial regression, **GAM uses smoothing splines to provide a flexible, non-parametric fit that better captures nonlinear patterns without imposing a global functional form**. The **best model** employed smoothing splines for $\log(\text{wage})$ and $\text{age}$, with smooth functions varying by marital status (referred to as gam12 in the internal analysis):
 
$$\text{Resp} \sim s(\text{logwage}, \text{by} = \text{maritl}) + s(\text{age}, \text{by} = \text{maritl}) + \text{All Categoricals}$$
```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Assumes mgcv is loaded: library(mgcv)

# cat_vars: character vector of categorical variables to consider
# train, test: your train/test data frames (e.g., Wage.train, Wage.test)

test_gam_subsets <- function(train, test, cat_vars) {
  
  # Base GAM structure (adjust if you want no age interaction, etc.)
  base_str <- "Resp ~ s(logwage, by = maritl) + maritl + s(age, by = maritl)"
  
  # Generate all subsets of cat_vars (including empty set)
  all_subsets <- unlist(
    lapply(0:length(cat_vars), function(k) {
      if (k == 0) {
        list(character(0))   # empty subset
      } else {
        combn(cat_vars, k, simplify = FALSE)
      }
    }),
    recursive = FALSE
  )
  
  results <- data.frame()
  
  for (vars in all_subsets) {
    if (length(vars) == 0) {
      # Base GAM only
      formula_str <- base_str
      label <- "Base: s(logwage, by=maritl) + maritl + s(age, by=maritl)"
    } else {
      added <- paste(vars, collapse = " + ")
      formula_str <- paste(base_str, "+", added)
      label <- paste0("Base + ", added)
    }
    
    fml <- as.formula(formula_str)
    
    # Fit GAM
    fit <- gam(fml, data = train)
    
    # Predict & compute Test MSE
    preds <- predict(fit, newdata = test)
    mse   <- mean((test$Resp - preds)^2)
    
    results <- rbind(
      results,
      data.frame(
        Model        = label,
        Test_MSE     = mse,
        stringsAsFactors = FALSE
      )
    )
  }
  
  # Sort by Test MSE (best first)
  results[order(results$Test_MSE), ]
}
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
cat_vars <- c("education", "race", "jobclass", "health", "health_ins")

gam_subset_results <- test_gam_subsets(
  train = Wage.train,
  test  = Wage.test,
  cat_vars = cat_vars
)

gam_subset_results %>%
  arrange(Test_MSE) %>%
  slice(1:5) %>%
  kable(
    caption = "Top 5 GAM Models by Test MSE",
    digits = 3
  ) %>%
  kable_styling(full_width = FALSE, font_size = 7)
```


Similar to polynomial regression, these results suggest that the GAM framework benefits from a full or nearly full list of additional categorical predictors, which enhance predictive accuracy when paired with spline-based nonlinear effects.

### Comparison to Polynomial Regression

Compared to the polynomial regression models, the GAM produced slightly higher Test MSE, likely because its smoother spline-based functions allows for flexibility by attempting to capture curvature in the data. This slight decrease in bias does result in a slightly lower residual variance, as observed in Table 8. The best polynomial model achieves slightly lower test MSE (0.976 vs 0.986) and lower 10-fold CV MSE (1.031 vs 1.056) than the GAM, indicating marginally better out-of-sample performance and stability with fewer effective degrees of freedom (22 vs 31.11). Although the GAM attains a very slightly lower residual variance, this comes at the cost of greater complexity, so the polynomial model offers a better balance of accuracy, parsimony, and interpretability.

```{r, echo = FALSE, include = FALSE, warning = FALSE, message = FALSE}
# Best GAM with all categorical predictors
gam_best_full <- gam(
  Resp ~ 
    s(logwage, by = maritl) +      # smooth of logwage by marital status
    maritl +                       # main effect of marital status
    s(age, by = maritl) +          # smooth of age by marital status
    education + race + jobclass +  # demographic / occupational covariates
    health + health_ins,           # health-related covariates
  data = Wage.train
)

# Test set predictions and Test MSE
pred_gam_best_full <- predict(gam_best_full, newdata = Wage.test)
mse_gam_best_full  <- mean((pred_gam_best_full - Wage.test$Resp)^2)
```

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}

# Distinguish best models for comparison
best_poly <- poly_full_all
best_gam  <- gam_best_full     

# Residual Variance
var_poly <- var(residuals(best_poly))
var_gam  <- var(residuals(best_gam))

variance_compare <- data.frame(
  Model = c("Best Polynomial", "Best GAM"),
  Residual_Variance = c(var_poly, var_gam)
)

variance_compare %>%
  kable(caption = "Residual Variance Comparison") %>%
  kable_styling(full_width = FALSE, font_size = 7)

```

```{r, echo = FALSE, warning = FALSE, message = FALSE, include = FALSE}

# ---- POLYNOMIAL DF ----
n_poly <- nobs(best_poly)
p_poly <- length(coef(best_poly))
resid_df_poly <- df.residual(best_poly)

# ---- GAM DF ----
gam_sum <- summary(best_gam)
param_df_gam <- length(gam_sum$p.coeff)
smooth_edf_sum <- sum(gam_sum$edf)
total_edf_gam <- param_df_gam + smooth_edf_sum
resid_df_gam <- best_gam$df.residual

# ---- Build DF Table ----
df_compare <- data.frame(
  Model = c("Best Polynomial", "Best GAM"),
  `Parametric DF` = c(p_poly, param_df_gam),
  `Smooth EDF` = c(0, round(smooth_edf_sum, 2)),
  `Total Effective DF` = c(p_poly, round(total_edf_gam, 2)),
  `Residual DF` = c(resid_df_poly, resid_df_gam)
)

# ---- Fancy Table ----
kable(df_compare,
      caption = "Comparison of Degrees of Freedom: Polynomial vs. GAM",
      digits = 2,
      align = "c") %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 7)

```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Cross Validation for Polynomial
poly_cv <- function(formula, data, k = 10) {
  set.seed(4620)  
  folds <- sample(rep(1:k, length.out = nrow(data)))
  cv_errors <- numeric(k)
  
  for (i in 1:k) {
    # Train/test split for this fold
    train_data <- droplevels(data[folds != i, ])
    test_data  <- droplevels(data[folds == i, ])
    
    # Fit polynomial model
    model <- lm(formula, data = train_data)
    
    # Predict and compute fold MSE
    preds <- predict(model, newdata = test_data)
    cv_errors[i] <- mean((test_data$Resp - preds)^2)
  }
  
  mean(cv_errors)
}


poly_cv_val <- poly_cv(
  Resp ~ poly(logwage, 3) * maritl +
         poly(age, 2) * maritl +
         education + race + jobclass + health + health_ins,
  data = Wage.train
)


# Cross Validation for GAM
gam_cv <- gam_cv <- function(model_formula, data, k = 10) {
  folds <- sample(rep(1:k, length.out = nrow(data)))
  cv_errors <- numeric(k)

  for (i in 1:k) {
    train_data <- data[folds != i, ]
    test_data  <- data[folds == i, ]
    
    model <- gam(model_formula, data = train_data)
    preds <- predict(model, newdata = test_data)
    cv_errors[i] <- mean((test_data$Resp - preds)^2)
  }
  mean(cv_errors)
}

gam_cv_val <- gam_cv(
  Resp ~ s(logwage, by = maritl) + maritl + s(age, by = maritl) +
    education + race + jobclass + health + health_ins,
  data = Wage.train
)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
cv_compare <- data.frame(
  Model = c("Best Polynomial", "Best GAM"),
  CV_MSE = c(poly_cv_val, gam_cv_val)
)

cv_compare %>%
  kable(caption = "Cross Validation Comparison") %>%
  kable_styling(full_width = FALSE, font_size = 7)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Side-by-side comparison of key metrics
model_compare <- data.frame(
  Metric = c("Test MSE",
             "10-fold CV MSE",
             "Residual Variance",
             "Total Effective DF"),
  `Best Polynomial` = c(
    round(mse_poly_full_all, 3),
    round(poly_cv_val, 3),
    round(var_poly, 3),
    round(p_poly, 2)
  ),
  `Best GAM` = c(
    round(mse_gam_best_full, 3),
    round(gam_cv_val, 3),
    round(var_gam, 3),
    round(total_edf_gam, 2)
  )
)

model_compare %>%
  kable(caption = "Comparison of Best Polynomial and GAM Models") %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 7)
```


## 4.3 Tree Based Method

Tree-based approaches were implemented to compare the performance of parametric models against nonparametric methods. CART, bagging, random forests, and boosting were fitted to the data, allowing for flexible modeling of the response variable.

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
# Ensure consistent factor types/levels for tree methods
factor_vars <- c("race", "education", "region", "jobclass", "health", "health_ins", "maritl")

# Make factors in TRAIN
Wage.train <- Wage.train %>%
  mutate(across(all_of(factor_vars), as.factor))

# Make factors in TEST with levels of TRAIN
Wage.test <- Wage.test %>%
  mutate(
    race       = factor(race,       levels = levels(Wage.train$race)),
    education  = factor(education,  levels = levels(Wage.train$education)),
    region     = factor(region,     levels = levels(Wage.train$region)),
    jobclass   = factor(jobclass,   levels = levels(Wage.train$jobclass)),
    health     = factor(health,     levels = levels(Wage.train$health)),
    health_ins = factor(health_ins, levels = levels(Wage.train$health_ins)),
    maritl     = factor(maritl,     levels = levels(Wage.train$maritl))
  )

```




```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
#### CART
# Fit full regression tree on training data
tree_fit <- tree(Resp ~ . -wage, data = Wage.train)
summary(tree_fit)
```


```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
# Plot full tree
plot(tree_fit, type = "uniform")
text(tree_fit, pretty = 0, cex = 0.5)
```



```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
#### Prune Tree
# Cross-validation to choose tree size
cv_tree <- cv.tree(tree_fit)     # uses deviance 
best_size <- cv_tree$size[which.min(cv_tree$dev)]

# Prune to optimal size
prune_fit <- prune.tree(tree_fit, best = best_size)

plot(prune_fit)
text(prune_fit, pretty = 0, cex = 0.7)
```

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
# Test MSE for pruned tree
tree_pred <- predict(prune_fit, newdata = Wage.test)
tree_mse  <- mean((tree_pred - Wage.test$Resp)^2)
tree_mse
```

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
# Bagging

p <- ncol(Wage.train) - 1  # number of predictors (exclude Resp)

bag_fit <- randomForest(
  Resp ~ .,
  data       = Wage.train,
  subset     = NULL,
  mtry       = p,          # bagging: use ALL predictors at each split
  importance = TRUE,
  ntree      = 500         # like Boston example
)

bag_pred <- predict(bag_fit, newdata = Wage.test)
bag_mse  <- mean((bag_pred - Wage.test$Resp)^2)
bag_mse

# Variable importance
varImpPlot(bag_fit, type = 1, main = "Bagging: Variable Importance")
```


```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
# Random Forest
rf_fit <- randomForest(
  Resp ~ .,
  data       = Wage.train,
  importance = TRUE          # so we can plot importance
  # mtry left as default: p/3 for regression
)

rf_pred <- predict(rf_fit, newdata = Wage.test)
rf_mse  <- mean((rf_pred - Wage.test$Resp)^2)
rf_mse

# Variable importance plot (like notes)
par(mfrow = c(1,1))
varImpPlot(rf_fit, type = 1, main = "Random Forest: Variable Importance")
```



```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
# Boosting

Wage.train_model <- Wage.train
Wage.test_model  <- Wage.test
Wage.train_model$maritl_original <- NULL
Wage.test_model$maritl_original  <- NULL
```


```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
boost_fit <- gbm(
  Resp ~ .,
  data              = Wage.train_model,
  distribution      = "gaussian",
  n.trees           = 5000,        # more trees; CV will pick the best
  interaction.depth = 2,
  shrinkage         = 0.01,        # typical learning rate
  cv.folds          = 10,          # 10-fold CV
  n.cores           = 1,           # or parallel if you want
  verbose           = FALSE
)
```


```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
### Getting the CV MSE, Residual Variance, and effective degrees of freedom for all the tree based methods and putting them into one table
## --- CART (pruned tree) ---

cv_tree   <- cv.tree(tree_fit)     
best_size <- cv_tree$size[which.min(cv_tree$dev)]
prune_fit <- prune.tree(tree_fit, best = best_size)

# Test MSE
cart_test_mse <- mean((predict(prune_fit, newdata = Wage.test) - Wage.test$Resp)^2)

# CV MSE: deviance / n as approximation
cart_cv_mse <- min(cv_tree$dev) / nrow(Wage.train)

# Residual variance on TRAIN
cart_resid  <- Wage.train$Resp - predict(prune_fit, newdata = Wage.train)
cart_resvar <- var(cart_resid)

# Effective DF = number of terminal nodes
cart_edf <- length(unique(prune_fit$where))


## --- Bagging (randomForest with mtry = p) ---

p <- ncol(Wage.train) - 1  # number of predictors

# bag_fit was already fit above
bag_test_mse <- mean((predict(bag_fit, newdata = Wage.test) - Wage.test$Resp)^2)

# OOB MSE as CV MSE
bag_cv_mse <- bag_fit$mse[bag_fit$ntree]

# Residual variance on TRAIN
bag_resid  <- Wage.train$Resp - predict(bag_fit, newdata = Wage.train)
bag_resvar <- var(bag_resid)

# EDF proxy from variance reduction
train_var  <- var(Wage.train$Resp)
bag_edf    <- nrow(Wage.train) * (1 - bag_cv_mse / train_var)


## --- Random Forest ---

rf_test_mse <- mean((predict(rf_fit, newdata = Wage.test) - Wage.test$Resp)^2)

# OOB MSE
rf_cv_mse <- rf_fit$mse[rf_fit$ntree]

# Residual variance on TRAIN
rf_resid  <- Wage.train$Resp - predict(rf_fit, newdata = Wage.train)
rf_resvar <- var(rf_resid)

# EDF proxy
rf_edf <- nrow(Wage.train) * (1 - rf_cv_mse / train_var)


## --- Boosting ---

# Find best number of trees by CV
best_iter <- gbm.perf(boost_fit, method = "cv", plot.it = FALSE)

best_iter   # you can print this to see

# CV MSE (average CV loss at best_iter)
boost_cv_mse <- boost_fit$cv.error[best_iter]

# Test MSE using best_iter
boost_pred     <- predict(boost_fit, newdata = Wage.test_model, n.trees = best_iter)
boost_test_mse <- mean((boost_pred - Wage.test_model$Resp)^2)

# Residual variance on TRAIN
boost_resid  <- Wage.train_model$Resp -
  predict(boost_fit, newdata = Wage.train_model, n.trees = best_iter)
boost_resvar <- var(boost_resid)

# EDF proxy
boost_edf <- best_iter * 2  # interaction.depth = 2


## --- Final table of tree-based methods ---

tree_results_full <- data.frame(
  Model             = c("CART (Pruned)", "Bagging", "Random Forest", "Boosting"),
  Test_MSE          = c(cart_test_mse,   bag_test_mse, rf_test_mse,   boost_test_mse),
  CV_MSE            = c(cart_cv_mse,     bag_cv_mse,   rf_cv_mse,     boost_cv_mse),
  Residual_Variance = c(cart_resvar,     bag_resvar,   rf_resvar,     boost_resvar),
  Effective_DF      = c(cart_edf,        bag_edf,      rf_edf,        boost_edf)
)

tree_results_full[order(tree_results_full$Test_MSE), ]
```

```{r, echo = FALSE, warning = FALSE}
# Your tree_results_full data frame (from earlier)
tree_results_full <- data.frame(
  Model             = c("CART (Pruned)", "Bagging", "Random Forest", "Boosting"),
  Test_MSE          = c(3.527937, 1.506307, 1.676172, 1.390001),
  CV_MSE            = c(4.213936, 1.577731, 1.583218, 1.564685),
  Residual_Variance = c(3.4277357, 0.3431415, 0.4757296, 1.2120234),
  Effective_DF      = c(8.000, 2234.716, 2234.308, 9990.000)
)

# Round values for presentation
tree_results_rounded <- tree_results_full %>%
  mutate(across(c(Test_MSE, CV_MSE, Residual_Variance, Effective_DF), ~ round(., 3)))

# Create a nice kable table
kable(tree_results_rounded, 
      col.names = c("Model", "Test MSE", "CV MSE", "Residual Variance", "Effective DF"),
      booktabs = TRUE, 
      digits = 3,
      caption = "Comparison of Tree Based Methods") %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 7)
```
Looking at the table above, we can see that Boosting has the lowest Test MSE, but not the lowest CV MSE, residual variance, or effective degrees of freedom.

- **CV MSE:** **Boosting** has the **lowest (1.565)**, which is better than Bagging (1.578) and Random Forest (1.583).
- **Residual Variance:** **Bagging** and **Random Forest** have much lower values (0.343 and 0.476), indicating they fit the training data better than Boosting (1.212).
- **Effective Degrees of Freedom:** **CART (Pruned)** is by far the simplest, with only **8.000**, while **Boosting** is the most complex **(9990.000)**.

So, **Boosting** is best for predicting on new data (lowest Test MSE), but it is more complex and has higher training error than Bagging and Random Forest. The low residual variance and CV MSE for Bagging and Random Forest suggest they may be more stable and less prone to overfitting, but they don't generalize quite as well to the test set as Boosting in this case. CART, although by far the simplest, performed the worst among the four tree-based methods. 

### Pros and Cons of Tree Based methods:

**Pros:** Tree ensembles can automatically capture nonlinear and complex interactions without explicitly specifying polynomial terms or interactions, which is useful given the interplay between marital status, age, and log(wage) observed in the exploratory analysis. They handle mixed numeric and categorical predictors naturally and provide variable-importance summaries to identify key features driving predictions.

**Cons:** Despite this flexibility, tree-based models do not achieve the low Test MSE of the best polynomial model or the competitive performance of the GAM, offering no predictive advantage. Additionally, their effective degrees of freedom are vastly higher than those of the polynomial and GAM models, reflecting a far more complex fit without meaningful improvement in accuracy.

# 5. Conclusion and Best Model Selection

The final model selection is based on a rigorous comparison of the best-performing models across three families, evaluating both predictive accuracy (Test MSE/CV MSE) and model complexity/stability (Effective Degrees of Freedom and Residual Variance).

```{r, echo = FALSE, message = FALSE, warning = FALSE}
best_models <- data.frame(
  Model = c("Best Polynomial",
            "Best GAM",
            "Best Tree (Boosting)"),
  Test_MSE = c(
    round(mse_poly_full_all, 3),
    round(mse_gam_best_full, 3),
    round(boost_test_mse, 3)
  ),
  check.names = FALSE,
  "10 fold CV MSE" = c(
    round(poly_cv_val, 3),
    round(gam_cv_val, 3),
    round(boost_cv_mse, 3)
  ),
  Residual_Variance = c(
    round(var_poly, 3),
    round(var_gam, 3),
    round(boost_resvar, 3)
  ),
  `Effective DF` = c(
    round(p_poly, 2),
    round(total_edf_gam, 2),
    round(boost_edf, 0)
  )
)

best_models %>%
  kable(
    caption = "Summary of Best Polynomial, GAM, and Tree-Based Models",
    booktabs = TRUE
  ) %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 7)

```

**Analysis of Model Performance and Trade-offs:**

**1. Best Polynomial Model:**

The best polynomial model shows the strongest predictive performance, with the **lowest Test MSE (0.976)** and **10-fold CV MSE (1.031)**, indicating accuracy and stability. It balances fit and complexity well, with low residual variance (1.013) and the **lowest Effective Degrees of Freedom (22.00)**, reflecting low bias and high parsimony. Its concise algebraic form also makes it the most interpretable, making it the preferred choice for statistical inference and clear communication.

**2. Generalized Additive Model (GAM):**

The GAM has slightly higher Test MSE (**0.986**) and CV MSE (**1.056**) than the polynomial model. Its smoothing functions offer more flexibility, giving a slightly lower residual variance (**1.010**) and better fit to the training data. However, this comes with greater complexity (Effective DF = **31.11**) without improving generalization, resulting in minor bias reduction but increased variance, making it a less optimal choice for prediction and practical use.

**3. Best Tree-Based Model (Boosting):**

The tree-based boosting model performs worse than the other approaches, with a Test MSE of **1.390**. It is highly complex (Effective DF = **9990.00**) and unstable, as shown by the large gap between its Test MSE and CV MSE (**1.565**), indicating severe overfitting. Its complexity and poor interpretability make it unsuitable for reliable prediction or meaningful statistical insights.


**Final Conclusion:**

The **Best Polynomial Model** remains the superior choice for this dataset.

It clearly and demonstrably outperforms all other models in the primary metrics of interest (lowest Test MSE and lowest CV MSE) while maintaining the lowest Effective DF. The Polynomial Model's combination of high accuracy, stability, and interpretability offers the best overall solution, minimizing the risk of over-fitting and maximizing the potential for clear inference.


\newpage

# Appendix


**Appendix 1**

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=3, fig.height=2.5}
numeric_vars <- data_clean %>% select(where(is.numeric))
corr_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")
corrplot(corr_matrix, 
         method = "color", 
         addCoef.col = "black", 
         tl.cex = 0.6,    
         number.cex = 0.5, 
         tl.srt = 45 
)
```

**Appendix 2**

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=3}
numeric_cols <- names(data_clean)[sapply(data_clean, is.numeric)]
pairs(data_clean[, numeric_cols], pch = 20, cex = 0.3, col = "darkblue")
```

**Appendix 3**

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}


cat_vars <- c("education", "race", "maritl", "jobclass", "health", "health_ins")

plot_list <- lapply(cat_vars, function(v) {
  ggplot(data_clean, aes_string(x = v, y = "Resp")) +
    geom_boxplot(fill = "grey") +
    labs(title = v, x = v, y = "Resp") +
    theme_minimal() +
    theme(plot.title = element_text(size = 6),
          axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
          axis.text.y = element_text(size = 5),
          axis.title = element_text(size = 6))
})
wrap_plots(plot_list, nrow = 2, ncol = 3)
```
