---
title: "STAT 4620 Project — Exploratory Data Analysis"
author: "Daphne Kaur (kaur.443)"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Introduction
This document provides an extended exploratory data analysis (EDA) for the Wage dataset.  

```{r}
load("Wage_Stat4620_2023.RData")
Wage <- Wage_Stat4620
head(Wage)
```

# 2. Understanding the Data

## 2.1 Dimensions, Structure, and Variable Types
```{r}
dim(Wage)
str(Wage)
sapply(Wage, class)
```

## 2.2 Missing Data Overview
```{r}
colSums(is.na(Wage))
mean(is.na(Wage))  # overall missing proportion
```

If missing values exist, inspect where they occur:
```{r}
which(is.na(Wage), arr.ind = TRUE)[1:20, ]
```

# 3. Summary Statistics

## 3.1 Numerical Variables
```{r}
numeric_vars <- names(Wage)[sapply(Wage, is.numeric)]
summary(Wage[, numeric_vars])
```

# 4. Exploratory Visualizations

## 4.1 Histograms
```{r}
par(mfrow=c(2,3))
for(v in numeric_vars){
  hist(Wage[[v]], main=paste("Histogram of", v), xlab=v, col="lightblue")
}
par(mfrow=c(1,1))
```

## 4.2 Density Plots
```{r}
par(mfrow=c(2,3))
for(v in numeric_vars){
  plot(density(Wage[[v]], na.rm=TRUE), main=paste("Density of", v))
}
par(mfrow=c(1,1))
```

## 4.3 Boxplots for Outliers
```{r}
par(mfrow=c(2,3))
for(v in numeric_vars){
  boxplot(Wage[[v]], main=paste("Boxplot of", v), col="tan")
}
par(mfrow=c(1,1))
```

## 4.4 Pairwise Relationships (Scatterplot Matrix)
```{r}
pairs(Wage[, numeric_vars], pch=20)
```

## 4.5 Correlation Matrix
```{r}
cor_matrix <- cor(Wage[, numeric_vars], use="complete.obs")
cor_matrix
```

# 5. Response Variable Exploration (Resp)

## 5.1 Basic Distribution
```{r}
hist(Wage$Resp, main="Distribution of Resp", xlab="Resp", col="lightgreen")
summary(Wage$Resp)
```

## 5.2 Resp vs Each Numerical Predictor
```{r}
par(mfrow=c(2,3))
for(v in numeric_vars){
  plot(Wage[[v]], Wage$Resp, pch=19, col="darkblue", 
       main=paste(v, "vs Resp"), xlab=v, ylab="Resp")
}
par(mfrow=c(1,1))
```

# 6. Checking Linearity (for regression models)
```{r}
par(mfrow=c(2,3))
for(v in numeric_vars){
  plot(Wage[[v]], Wage$Resp, pch=20, col="black")
  abline(lm(Resp ~ Wage[[v]], data=Wage), col="red", lwd=2)
}
par(mfrow=c(1,1))
```



