---
title: "Arjun Narang Final Project Modeling"
author: "Arjun Narang.37"
date: "2025-11-19"
output: pdf_document
---

### Load Packages and Data

```{r}
# Packages
if (!require("splines")) {
  install.packages("splines")
  library(splines)
}

if (!require("ISLR2")) {
  install.packages("ISLR2")
  library(ISLR2)
}

if (!require("dplyr")) {
  install.packages("dplyr")
  library(dplyr)
}

if (!require("corrplot")) {
  install.packages("corrplot")
  library(corrplot)
}

if (!require("ggplot2")) {
  install.packages("ggplot2")
  library(ggplot2)
}

if (!require("patchwork")) {
  install.packages("patchwork")
  library(patchwork)
}

if (!require("MASS")) {
  install.packages("MASS")
  library(MASS)
}

if (!require("tibble")) {
  install.packages("tibble")
  library(tibble)
}

if (!require("boot")) {
  install.packages("boot")
  library(boot)
}

if (!require("mgcv")) {
  install.packages("mgcv")
  library(mgcv)
}

if (!require("car")) {
  install.packages("car")
  library(car)
}

if (!require("tree")) {
  install.packages("tree")
  library(tree)
}

if (!require("randomForest")) {  # for bagging + RF
  install.packages("randomForest")
  library(randomForest)
}

if (!require("gbm")) {           # for boosting
  install.packages("gbm")
  library(gbm)
}


```



```{r}
load("Wage_Stat4620_2023.RData")
data <- Wage_Stat4620
```

### Clean Data

```{r}
# Get rid of all the N/A response variables
data_clean <- data %>% filter(!is.na(Resp))
data <- data_clean
dim(data)
```

```{r}
# Change marital status to married vs not married
data$maritl <- ifelse(data$maritl == "2. Married",
                      "Married",
                      "Not Married")

data$maritl <- factor(data$maritl,
                      levels = c("Not Married", "Married"))
```


We aggregate some of the values for maritl since we see a difference in married vs all the other values in the response variable. This will be observed below.

### EDA Views

```{r}
numeric_vars <- names(data)[sapply(data, is.numeric)]
par(mfrow=c(2,3))
for(v in numeric_vars){
  plot(data[[v]], data$Resp, pch=19, col="darkblue", 
       main=paste(v, "vs Resp"), xlab=v, ylab="Resp")
}
par(mfrow=c(1,1))
```



With some preliminary EDA, we can see that the response variable has a bi-modal distribution. Additionally it is slightly skewed to the right. This lack of normality in the distribution would violate most linear models. This means our modeling will be focusing in on models that allow interaction effects like polynomial model, GAMs, and possibly trees.

```{r}
# Categorical Predictors vs Response
cat_vars <- c("education", "race", "maritl", "jobclass", "health", "health_ins")

plot_list <- lapply(cat_vars, function(v) {
  ggplot(data, aes_string(x = v, y = "Resp")) +
    geom_boxplot(fill = "grey") +
    labs(title = paste("Resp by", v), x = v, y = "Resp") +
    theme_minimal()
})

combined_plot <- plot_list[[1]] + plot_list[[2]] + plot_list[[3]] +
                 plot_list[[4]] + plot_list[[5]] + plot_list[[6]] +
                 plot_layout(nrow = 3, ncol = 2)
combined_plot
```

It seems that the response does vary a bit amongst education, race, and marital status. These will likely be useful in helping us model the response variable. 





```{r}
pairs(data[, numeric_vars], pch=20)
```

Based off of the this we see that logwage and and age seem to be good predictors let's try to break them down further.


```{r}
plot_cat <- function(var) {
  ggplot(data, aes(x = logwage, y = Resp, color = .data[[var]])) +
    geom_point(alpha = 0.05, size = 0.4) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
    labs(title = paste("Resp vs logwage by", var),
         x = "log(wage)", y = "Resp") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")  # hides legends to shrink figure
}

p1 <- plot_cat("education")
p2 <- plot_cat("race")
p3 <- plot_cat("maritl")
p4 <- plot_cat("jobclass")
p5 <- plot_cat("health")
p6 <- plot_cat("health_ins")

# Combine into a 3x2 grid
final_fig <- (p1 | p2 | p3) /
             (p4 | p5 | p6)

final_fig
```

```{r}
# Create each plot
plot_age_cat <- function(var) {
  ggplot(data, aes(x = age, y = Resp, color = .data[[var]])) +
    geom_point(alpha = 0.05, size = 0.4) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
    labs(title = paste("Resp vs Age by", var),
         x = "Age", y = "Resp") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "none")
}

# Create each individual plot
p1 <- plot_age_cat("education")
p2 <- plot_age_cat("race")
p3 <- plot_age_cat("maritl")
p4 <- plot_age_cat("jobclass")
p5 <- plot_age_cat("health")
p6 <- plot_age_cat("health_ins")

# Combine into 3x2 patchwork layout
age_fig <- (p1 | p2 | p3) /
           (p4 | p5 | p6)

age_fig
```

Out of all the categorical variables, marital status seems to be the only clear interaction with logWage and age. When modeling we will definitely include that as an interaction and perhaps include education and race as a main effect and see how it performs with the test MSE.


### Test/Train Split

```{r}
set.seed(4620)

n <- nrow(data)
train_size <- round(n * 0.8) # Create an 80/20 split
train <- sample(1:n,train_size, replace = FALSE)
test <- -train

Wage.train <- data[train, ]
Wage.test  <- data[test, ]

```


### Modeling logWage against Response


```{r}
# Fit 3rd degree polynomial using basic, interaction, and GAM
fit_poly_3 <- lm(Resp ~ poly(logwage, 3), data = Wage.train)
fit_poly3_int <- lm(Resp ~ poly(logwage, 3) * maritl, data = Wage.train)
fit_gam_3 <- gam(Resp ~ s(logwage, by = maritl) + maritl, data = Wage.train)

summary(fit_poly3_int)
```


```{r}
res <- residuals(fit_poly3_int)

qqnorm(res, main = "Q-Q Plot of Residuals (Polynomial Degree 3)")
qqline(res, col = "red", lwd = 2)
```

We see the response being modeled quite well by the ploynomial, a bit of variability towards the ends of the plot, but we can work to imrpove that as we continue to evaluate variables.


```{r}
# Degree 3 predictions + MSE
pred3 <- predict(fit_poly_3, newdata = data[test, ])
pred3_int <- predict(fit_poly3_int, newdata = data[test, ]) 
pred_gam_3 <- predict(fit_gam_3, newdata = data[test, ])

# Calculate the MSE prayer
mse3  <- mean((pred3 - data[test, ]$Resp)^2)
mse3_int <- mean((pred3_int - data[test, ]$Resp)^2)
mse3_gam <- mean((pred_gam_3 - data[test, ]$Resp)^2)

mse_results <- data.frame(
  Model = c("Polynomial (Degree 3)", "Polynomial (Degree 3) with Marriage Interaction", " GAM"),
  Test_MSE = c(mse3, mse3_int, mse3_gam)
)

mse_results
```

We see that polynomial (degree 3) with interaction effects has the lowest test MSE. 

```{r}
# Add predictions to test set
Wage.test$pred_poly3_int <- pred3_int

ggplot(Wage.test, aes(x = logwage, y = Resp, color = maritl)) +
  geom_point(alpha = 0.1, size = 0.7) +
  geom_line(aes(y = pred_poly3_int),
            linewidth = 1.4) +
  labs(title = "Polynomial Degree 3 with Interaction: Resp vs logwage",
       x = "log(wage)", y = "Resp") +
  theme_minimal()

```

### Modeling Age vs Resp

```{r}
ggplot(data, aes(x = age, y = Resp, color = maritl)) +
  geom_point(alpha = 0.1, size = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.5) +
  labs(title = "Resp vs Age by Marital Status",
       x = "Age", y = "Resp") +
  theme_minimal()
```


```{r}
# Fit 2nd-degree polynomial using basic, interaction, and GAM for age
fit_age_poly2     <- lm(Resp ~ poly(age, 2), data = Wage.train)
fit_age_poly2_int <- lm(Resp ~ poly(age, 2) * maritl, data = Wage.train)
fit_age_gam       <- gam(Resp ~ s(age, by = maritl) + maritl, data = Wage.train)

summary(fit_age_poly2_int)
```

```{r}
res_age2 <- residuals(fit_age_gam)

qqnorm(res_age2, main = "Q-Q Plot of Residuals (Age Polynomial Degree 2 with Interaction)")
qqline(res_age2, col = "red", lwd = 2)
```

We see the ends of the Q-Q plot being modeled very poorly. This is likely due to the upper elements in Age around 40-50 being underestimated.


```{r}
# Predictions + MSE for age-based models
pred_age2      <- predict(fit_age_poly2,     newdata = Wage.test)
pred_age2_int  <- predict(fit_age_poly2_int, newdata = Wage.test)
pred_age_gam   <- predict(fit_age_gam,       newdata = Wage.test)

mse_age2      <- mean((pred_age2     - Wage.test$Resp)^2)
mse_age2_int  <- mean((pred_age2_int - Wage.test$Resp)^2)
mse_age2_gam  <- mean((pred_age_gam  - Wage.test$Resp)^2)

mse_age_results <- data.frame(
  Model    = c("Age Polynomial (Degree 2)",
               "Age Polynomial (Degree 2) with Marriage Interaction",
               "Age GAM"),
  Test_MSE = c(mse_age2, mse_age2_int, mse_age2_gam)
)

mse_age_results

```

```{r}
Wage.test$pred_age_gam <- pred_age_gam

ggplot(Wage.test, aes(x = age, y = Resp, color = maritl)) +
  geom_point(alpha = 0.1, size = 0.7) +
  geom_line(aes(y = pred_age_gam), linewidth = 1.4) +
  labs(title = "GAM for Age: Resp vs Age",
       x = "Age", y = "Resp") +
  theme_minimal()
```


### Final GAM Model

Now that we modeled both logWage and age, we must combine them. The best strategy for this would be to use a GAM that allows for different variables/interactions to explain parts of the variability.


```{r}
# BASE GAM (no age interaction)
gam_base <- gam(
  Resp ~ 
    s(logwage, by = maritl) + 
    maritl + 
    s(age), 
  data = Wage.train
)

# BASE GAM (age interaction)
gam_full_base <- gam(
  Resp ~ 
    s(logwage, by = maritl) + 
    maritl + 
    s(age, by = maritl), 
  data = Wage.train
)

# BASE GAM (no Maritl)
gam_full_base_test <- gam(
  Resp ~ 
    s(logwage, by = maritl) + 
    s(age, by = maritl), 
  data = Wage.train
)

# Add education
gam_full_edu_ageint <- gam(
  Resp ~ s(logwage, by = maritl) + maritl + 
    s(age, by = maritl) +
    education,
  data = Wage.train
)

# Add race
gam_full_race_ageint <- gam(
  Resp ~ s(logwage, by = maritl) + maritl + 
    s(age, by = maritl) +
    race,
  data = Wage.train
)

# Add jobclass
gam_full_job_ageint <- gam(
  Resp ~ s(logwage, by = maritl) + maritl + 
    s(age, by = maritl) +
    jobclass,
  data = Wage.train
)

# Add education + race
gam_full_edu_race_ageint <- gam(
  Resp ~ s(logwage, by = maritl) + maritl + 
    s(age, by = maritl) +
    education + race,
  data = Wage.train
)

# Add education + jobclass
gam_full_edu_job_ageint <- gam(
  Resp ~ s(logwage, by = maritl) + maritl + 
    s(age, by = maritl) +
    education + jobclass,
  data = Wage.train
)

# Add race + jobclass
gam_full_race_job_ageint <- gam(
  Resp ~ s(logwage, by = maritl) + maritl + 
    s(age, by = maritl) +
    race + jobclass,
  data = Wage.train
)

# FULL GAM with age interaction + all categoricals
gam_full_age_interaction <- gam(
  Resp ~ s(logwage, by = maritl) + maritl +
    s(age, by = maritl) +
    education + race + jobclass,
  data = Wage.train
)

```


```{r}
# Put all GAMs in a named list (with age interaction)
gam_age_int_models <- list(
  "GAM base (no age interaction)" = gam_base,
  "GAM base" = gam_full_base,
  "GAM base (no maritl)" = gam_full_base_test,
  "GAM + education"  = gam_full_edu_ageint,
  "GAM + race"       = gam_full_race_ageint,
  "GAM + jobclass"   = gam_full_job_ageint,
  "GAM + edu + race" = gam_full_edu_race_ageint,
  "GAM + edu + job"  = gam_full_edu_job_ageint,
  "GAM + race + job" = gam_full_race_job_ageint,
  "GAM FULL"           = gam_full_age_interaction
)

# Compute MSE
gam_age_int_mse <- sapply(gam_age_int_models, function(mod) {
  pred <- predict(mod, newdata = Wage.test)
  mean((pred - Wage.test$Resp)^2)
})

gam_age_int_mse_df <- data.frame(
  Model    = names(gam_age_int_mse),
  Test_MSE = as.numeric(gam_age_int_mse),
  row.names = NULL
)

gam_age_int_mse_df
```


We see the best Test MSE comes from using all the variables and interactions for marital status and logWage and age.


### Cross Validation

Now lets use cross validation to ensure that the best model CV MSE is selected.

```{r}

gam_cv <- function(formula, data, k = 10) {
  folds <- sample(rep(1:k, length.out = nrow(data)))
  cv_errors <- numeric(k)
  
  for (i in 1:k) {
    train_data <- data[folds != i, ]
    test_data  <- data[folds == i, ]
    
    model <- gam(formula, data = train_data)
    preds <- predict(model, newdata = test_data)
    cv_errors[i] <- mean((test_data$Resp - preds)^2)
  }
  
  mean(cv_errors)
}
```


```{r}
cv_base        <- gam_cv(Resp ~ s(logwage, by = maritl) + maritl + s(age, by = maritl), data)
cv_job         <- gam_cv(Resp ~ s(logwage, by = maritl) + maritl + s(age, by = maritl) + jobclass, data)
cv_edu_job     <- gam_cv(Resp ~ s(logwage, by = maritl) + maritl + s(age, by = maritl) + education + jobclass, data)
cv_full        <- gam_cv(Resp ~ s(logwage, by = maritl) + maritl + s(age, by = maritl) + education + race + jobclass, data)
```


```{r}
cv_results <- data.frame(
  Model = c("GAM base",
            "GAM + jobclass",
            "GAM + edu + jobclass",
            "GAM FULL"),
  CV_MSE = c(cv_base, cv_job, cv_edu_job, cv_full)
)

cv_results
```

We see through 10-fold cross validation that the Cross Validation MSE is actually lowest when considering interactions with age/logWage with marital status and only including job class.

```{r}
# Extract residuals from the GAM
res_gam <- residuals(gam_full_job_ageint)

# Q–Q Plot
qqnorm(res_gam,
       main = "Q–Q Plot of Residuals\n(GAM: logwage×maritl + age×maritl + jobclass)",
       pch = 19, cex = 0.5)
qqline(res_gam, col = "red", lwd = 2)

```

That being said, we will still likely go with Test MSE as our metric of choice since CV MSE would be more useful in a hyperparameter tuning scenario, which this isn't.


```{r}
# Extract residuals from the GAM
res_gam <- residuals(gam_full_age_interaction)

# Q–Q Plot
qqnorm(res_gam,
       main = "Q–Q Plot of Residuals\n(GAM: logwage×maritl + age×maritl + jobclass + edu + race)",
       pch = 19, cex = 0.5)
qqline(res_gam, col = "red", lwd = 2)

```

### Trees, Random Forest, Bagging, and Boosting

#### Variable alteration for tree use

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
# Ensure consistent factor types/levels for tree methods
factor_vars <- c("race", "education", "region", "jobclass", "health", "health_ins", "maritl")

# Make factors in TRAIN
Wage.train <- Wage.train %>%
  mutate(across(all_of(factor_vars), as.factor))

# Make factors in TEST with levels of TRAIN
Wage.test <- Wage.test %>%
  mutate(
    race       = factor(race,       levels = levels(Wage.train$race)),
    education  = factor(education,  levels = levels(Wage.train$education)),
    region     = factor(region,     levels = levels(Wage.train$region)),
    jobclass   = factor(jobclass,   levels = levels(Wage.train$jobclass)),
    health     = factor(health,     levels = levels(Wage.train$health)),
    health_ins = factor(health_ins, levels = levels(Wage.train$health_ins)),
    maritl     = factor(maritl,     levels = levels(Wage.train$maritl))
  )

```


#### CART

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
# Fit full regression tree on training data
tree_fit <- tree(Resp ~ . -wage, data = Wage.train)
summary(tree_fit)
```


```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
# Plot full tree
plot(tree_fit, type = "uniform")
text(tree_fit, pretty = 0, cex = 0.5)
```

#### Prune Tree

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
# Cross-validation to choose tree size
cv_tree <- cv.tree(tree_fit)     # uses deviance 
best_size <- cv_tree$size[which.min(cv_tree$dev)]

# Prune to optimal size
prune_fit <- prune.tree(tree_fit, best = best_size)

plot(prune_fit)
text(prune_fit, pretty = 0, cex = 0.7)
```

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
# Test MSE for pruned tree
tree_pred <- predict(prune_fit, newdata = Wage.test)
tree_mse  <- mean((tree_pred - Wage.test$Resp)^2)
tree_mse
```

#### Bagging

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}

p <- ncol(Wage.train) - 1  # number of predictors (exclude Resp)

bag_fit <- randomForest(
  Resp ~ .,
  data       = Wage.train,
  subset     = NULL,
  mtry       = p,          # bagging: use ALL predictors at each split
  importance = TRUE,
  ntree      = 500         # like Boston example
)

bag_pred <- predict(bag_fit, newdata = Wage.test)
bag_mse  <- mean((bag_pred - Wage.test$Resp)^2)
bag_mse

# Variable importance
varImpPlot(bag_fit, type = 1, main = "Bagging: Variable Importance")
```

#### Random Forest

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
rf_fit <- randomForest(
  Resp ~ .,
  data       = Wage.train,
  importance = TRUE          # so we can plot importance
  # mtry left as default: p/3 for regression
)

rf_pred <- predict(rf_fit, newdata = Wage.test)
rf_mse  <- mean((rf_pred - Wage.test$Resp)^2)
rf_mse

# Variable importance plot (like notes)
par(mfrow = c(1,1))
varImpPlot(rf_fit, type = 1, main = "Random Forest: Variable Importance")
```

#### Boosting

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}

boost_fit <- gbm(
  Resp ~ .,
  data              = Wage.train,
  distribution      = "gaussian",
  n.trees           = 5000,        # more trees; CV will pick the best
  interaction.depth = 2,
  shrinkage         = 0.01,        # typical learning rate
  cv.folds          = 10,          # 10-fold CV
  n.cores           = 1,           # or parallel if you want
  verbose           = FALSE
)

```

### Getting the CV MSE, Residual Variance, and effective degrees of freedom for all the tree based methods and putting them into one table

```{r, echo = FALSE, include = FALSE, warning = FALSE, message=FALSE}
### Tree-based metrics: Test MSE, CV MSE, Residual Variance, EDF

## --- CART (pruned tree) ---

cv_tree   <- cv.tree(tree_fit)     
best_size <- cv_tree$size[which.min(cv_tree$dev)]
prune_fit <- prune.tree(tree_fit, best = best_size)

# Test MSE
cart_test_mse <- mean((predict(prune_fit, newdata = Wage.test) - Wage.test$Resp)^2)

# CV MSE: deviance / n as approximation
cart_cv_mse <- min(cv_tree$dev) / nrow(Wage.train)

# Residual variance on TRAIN
cart_resid  <- Wage.train$Resp - predict(prune_fit, newdata = Wage.train)
cart_resvar <- var(cart_resid)

# Effective DF = number of terminal nodes
cart_edf <- length(unique(prune_fit$where))


## --- Bagging (randomForest with mtry = p) ---

p <- ncol(Wage.train) - 1  # number of predictors

# bag_fit was already fit above
bag_test_mse <- mean((predict(bag_fit, newdata = Wage.test) - Wage.test$Resp)^2)

# OOB MSE as CV MSE
bag_cv_mse <- bag_fit$mse[bag_fit$ntree]

# Residual variance on TRAIN
bag_resid  <- Wage.train$Resp - predict(bag_fit, newdata = Wage.train)
bag_resvar <- var(bag_resid)

# EDF proxy from variance reduction
train_var  <- var(Wage.train$Resp)
bag_edf    <- nrow(Wage.train) * (1 - bag_cv_mse / train_var)


## --- Random Forest ---

rf_test_mse <- mean((predict(rf_fit, newdata = Wage.test) - Wage.test$Resp)^2)

# OOB MSE
rf_cv_mse <- rf_fit$mse[rf_fit$ntree]

# Residual variance on TRAIN
rf_resid  <- Wage.train$Resp - predict(rf_fit, newdata = Wage.train)
rf_resvar <- var(rf_resid)

# EDF proxy
rf_edf <- nrow(Wage.train) * (1 - rf_cv_mse / train_var)


## --- Boosting ---

# Find best number of trees by CV
best_iter <- gbm.perf(boost_fit, method = "cv", plot.it = FALSE)

best_iter   # you can print this to see

# CV MSE (average CV loss at best_iter)
boost_cv_mse <- boost_fit$cv.error[best_iter]

# Test MSE using best_iter
boost_pred     <- predict(boost_fit, newdata = Wage.test, n.trees = best_iter)
boost_test_mse <- mean((boost_pred - Wage.test$Resp)^2)

# Residual variance on TRAIN
boost_resid  <- Wage.train$Resp -
  predict(boost_fit, newdata = Wage.train, n.trees = best_iter)
boost_resvar <- var(boost_resid)

# EDF proxy
boost_edf <- best_iter * 2  # interaction.depth = 2


## --- Final table of tree-based methods ---

tree_results_full <- data.frame(
  Model             = c("CART (Pruned)", "Bagging", "Random Forest", "Boosting"),
  Test_MSE          = c(cart_test_mse,   bag_test_mse, rf_test_mse,   boost_test_mse),
  CV_MSE            = c(cart_cv_mse,     bag_cv_mse,   rf_cv_mse,     boost_cv_mse),
  Residual_Variance = c(cart_resvar,     bag_resvar,   rf_resvar,     boost_resvar),
  Effective_DF      = c(cart_edf,        bag_edf,      rf_edf,        boost_edf)
)

tree_results_full[order(tree_results_full$Test_MSE), ]


```


### Using Kable package for the table
```{r, echo = FALSE, warning = FALSE}
library(knitr)
library(kableExtra)

# Your tree_results_full data frame (from earlier)
tree_results_full <- data.frame(
  Model             = c("CART (Pruned)", "Bagging", "Random Forest", "Boosting"),
  Test_MSE          = c(3.527937, 1.471860, 1.603544, 1.419123),
  CV_MSE            = c(4.252760, 1.596224, 1.512284, 1.640975),
  Residual_Variance = c(3.4277357, 0.3190218, 0.3627335, 1.1903342),
  Effective_DF      = c(8.000, 2233.342, 2239.581, 9834.000)
)

# Round values for presentation
tree_results_rounded <- tree_results_full %>%
  mutate(across(c(Test_MSE, CV_MSE, Residual_Variance, Effective_DF), ~ round(., 3)))

# Create a nice kable table
kable(tree_results_rounded, 
      col.names = c("Model", "Test MSE", "CV MSE", "Residual Variance", "Effective DF"),
      booktabs = TRUE, 
      digits = 3) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

Looking at the table above, we can see that Boosting has the lowest Test MSE, but not the lowest CV MSE, residual variance, or effective degrees of freedom.

- CV MSE: Random Forest has the lowest (1.512), slightly better than Boosting (1.641).
- Residual Variance: Bagging and Random Forest have much lower values (0.319 and 0.363), indicating they fit the training data better than Boosting (1.190).
- Effective Degrees of Freedom: CART (Pruned) is by far the simplest, with only 8, while Boosting is the most complex (9834).

So, Boosting is best for predicting on new data (lowest Test MSE), but it is more complex and has higher training error than Bagging and Random Forest. The low residual variance and CV MSE for Bagging and Random Forest suggest they may be more stable and less prone to overfitting, but they don't generalize quite as well to the test set as Boosting in this case. CART, although by far the simplest, performed the worst among the four tree-based methods. 

### Pros of Tree Based methods:

Tree ensembles automatically capture nonlinear and complex interactions without requiring explicit specification of polynomial terms or interaction effects (like we have done for the GAMs and polynomial regression models), which is appealing because of the interaction between marital status, age, and log(wage) observed in the exploratory analysis. They also handle mixed numeric and categorical predictors naturally and give variable-importance summaries that can highlight which demographic and job-related features drive predictions.

### Cons of Tree Based methods:

However, despite this flexibility, none of the tree-based models are able to reach the very low test MSE achieved by the best polynomial model, nor the slightly higher but still competitive GAM, meaning they offer no predictive advantage here. In addition, the effective degrees of freedom for ensembles such as boosting are many orders of magnitude larger than those of the polynomial and GAM models, indicating a much more complex fit with little to show in terms of improved accuracy.

### Interpretability and stability:

The parametric models are also preferred due to their interpretability: the polynomial model derives a concise equation with a small number of coefficients, whereas boosted trees, for example, need to aggregate thousands of small trees whose combined effect is extremely difficult to explain qualitatively. Moreover, the GAM and polynomial fits exhibit stable behavior under cross-validation with relatively low CV MSE and residual variance, while the tree ensembles introduce additional tuning complexity (number of trees, depth, learning rate) without giving any clearly better bias–variance trade-offs.